{RILS}-{ROLS}: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-12-13 paper_resources/doc/bmc_jbd/bmc_jbd.tex
English revisions applied.

\title{ {RILS}-{ROLS} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares}

@@

\begin{abstract} % abstract
In this paper, we solve the well-known symbolic regression problem that has been intensively studied in the literature and has a wide range of applications. 
% ACA: ove recenice vise prilice uvodu, tamo svakako i postoji, da ne bude preglomazan abstrakt
%The symbolic regression models could provide insights towards establishing physical theory behind experimentally observed data. 
%Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology, modeling analytic representations of the exciton binding energy in energy science, finding models for band gaps of NaCl-type compounds in material science, just to name a few.

We propose an efficient meta-heuristic-based approach, called \textsc{RILS}-\textsc{ROLS} to solve it. \textsc{RILS}-\textsc{ROLS} is based on the following two parts: ($i$) iterated local search serving as the method backbone, mainly solving combinatorial and some continuous aspects of the problem; ($ii$) ordinary least square method is focused on the continuous aspect of the search space -- it efficiently determines the best--fitting coefficients of linear combinations within solution equations. Beside that, we introduce a novel fitness function that combines three important model quality measures: $R^2$ score, RMSE score, and the size of the model (or model complexity), and carefully designed local search, which allows systematic search in proximity to candidate solution.

The experiments are conducted on the two well-known ground-truth benchmark sets called \textsc{Feynman} and \textsc{Strogatz}. \textsc{RILS}-\textsc{ROLS} was compared to 14 other competitors from the literature. Our method outperformed all 14 competitors with respect to the true symbolic model accuracy under varying levels of noise. It also showed to be the most efficient considering the average running time of reaching the exact model.  
In addition to evaluation on known ground-truth datasets, a new randomly generated set of problem instances is introduced. The goal of \textsc{Random} dataset was to test the scalability of our method with respect to incremental equation sizes and number of variables, under different levels of noise.

Statistical analysis of obtained experimental results confirmed that \textsc{RILS}-\textsc{ROLS} can be considered as a new state-of-the-art method for solving the problem of symbolic regression when ground-truth equations are known.
\end{abstract}

@@

The problem of symbolic regression (SR)~\cite{billard2002symbolic} has attracted many researchers over the last decade. SR can be seen as a generalization of more specific variants of regression in which the functional form is fixed, e.g. the well-known linear regression, polynomial regression~\cite{stimson1978interpreting}, etc. All regression models have the same goal: given a set of $n$-dimensional input data and its corresponding continuous output target variable, the aim is to find a  mathematical expression (function) of $n$ (input) variables that   \emph{fits} best to the target variable.  %This computationally intensive task is in general provenly NP--hard~\cite{virgolin2022symbolic}. 
In the case of linear regression, the model is always a linear combination of input variables. This is in general not good enough, since target variable might be dependent on some nonlinear function among input variables. Unlike linear regression, SR allows the search over much larger space of possible mathematical formulas to find the best-fitting ones, i.e. those able to predict the target variable from the input variables. The basis of constructing explicit formula is in elementary operations like addition and multiplication, as well as polynomial, trigonometric, exponential, and others.  
%Application: https://towardsdatascience.com/real-world-applications-of-symbolic-regression-2025d17b88ef

SR models tend to be interpretable, unlike artificial neural networks. Coefficients inside SR formulas might indicate the absolute or relative importance of certain input variables. Moreover, relations among variables might be described in an elegant way. For example, the appearance of an exponential function can be associated with some physical phenomenon such as the intensity of radiation or acceleration over time, see~\cite{udrescu2020ai}. Additionally, the SR models often have high generalization power unlike some models with fixed functional forms such as polynomial regression.

Practical applications of SR in chemical and biological sciences are listed in~\cite{weng2020simple}. In particular, this paper describes the discovery of a series of new oxide perovskite catalysts with improved activities. The application of SR to discovering physical laws from a distorted video is studied in~\cite{udrescu2021symbolic}. Revealing complex ecological dynamics by SR is presented in~\cite{chen2019revealing}. Paper \cite{louis2021reviewing} provides the application of SR to model the effects of the mutations on protein stability, in the domain of fundamental and applied biology. A recent study in \cite{liu2021machine} is concerned of auto-discovering conserved quantities using trajectory data from unknown dynamical systems. The application of SR to model analytic representations of the exciton binding energy is shown in~\cite{liang2019phillips}. The use of SR in material science is described in~\cite{wang2019symbolic,wang2022symbolic,burlacu2022symbolic,kabliman2021application}. SR application to wind speed forecasting is given in~\cite{abdellaoui2021symbolic}.

	There are many different ways to tackle the SR. Most of them are based on   machine learning techniques, genetic programming (GP) or some other meta-heuristics. Among the first GP methods to tackle SR is the one of Raidl~\cite{raidl1998hybrid} based on a hybrid variant of genetic programming. A differential evolution algorithm was proposed by Cerny et al.~\cite{cerny2008using}. Age-fitness Pareto Optimization approach is proposed by Smidt and Lipson~\cite{schmidt2010age}.  Application of the artificial bee colony programming to solve SR is proposed by Karaboga et al.~\cite{karaboga2012artificial}. 
Application of local-based heuristics for solving SR is reported by Commenda in his PhD thesis~\cite{kommenda2018local}. A GP-based approach, the gene-pool optimal mixing evolutionary algorithm (\textsc{GOMEA}) is studied by Virgolin et al.~\cite{virgolin2021improving}.  Another evolutionary algorithm, the interaction-transformation EA (\textsc{ITEA}) has been proposed by de Franca et al.~\cite{de2021interaction}. Simulated annealing to solve SR is proposed by Kantor~\cite{kantor2021simulated}. A variable neighbourhood programming approach  is proposed by Elleurich et al.~\cite{elleuch2020variable}; this technique is initially proposed in~\cite{elleuch2016variable}. 
Kommenda et al.~\cite{kommenda2020parameter} proposed a method called \textsc{Operon} algorithm, which uses nonlinear least squares for parameter identification of SR models further integrated into a local search mechanism in tree-based GP. The C++ implementation of \textsc{Operon} is discussed in~\cite{burlacu2020operon}. The method that utilizes the Taylor polynomials to approximate the symbolic equation that fits the dataset, called Taylor genetic programming, is proposed in~\cite{he2022taylor}. A GP approach that uses the idea of semantic back-propagation (SBP-GP) is proposed in~\cite{virgolin2019linear}.  Empirical analysis of variance between many GP-based methods for SR is discussed in~\cite{kammerer2021empirical}. It is also worth to mention the GP-based Eurequa commercial solver~\cite{schmidt2009distilling, schmidt2011machine} that uses age-fitness pareto optimization with co-evolved fitness estimation. This solver is nowadays accepted as the gold standard of symbolic regression.

	Method based on Bayesian symbolic regression (\textsc{BSR}) is proposed in~\cite{jin2019bayesian} -- this method belongs to a family of Markov Chain Monte Carlo algorithms (MCMC). Deep Symbolic Regression (\textsc{DSR}), an RNN approach, is proposed in~\cite{petersen2019deep} -- it utilizes the policy gradient search. This mechanism of search is further investigated in~\cite{landajuela2021improving}. A fast neural network approach, called \textsc{OccamNet},  is proposed in~\cite{costa2020fast}.  A deep reinforcement learning approach enhanced with genetic programming is proposed in~\cite{mundhenk2021symbolic}.

Powerful hybrid techniques to solve SR are also well studied; among them, we emphasize the \textsc{Eplex} solver from~\cite{la2019probabilistic,la2016epsilon} and \textsc{AI-Feynman} algorithm from~\cite{udrescu2020ai}, a physics-inspired divide-and-conquer method combined with the neural network fitting. The latter is one of the most efficient methods for physically-inspired models. We also mention the Fast Function Extraction (\textsc{FFX}) algorithm developed by McConaghy~\cite{mcconaghy2011ffx}, which is a non-evolutionary method combined with a machine learning technique called path-wise-regularized learning, which quickly prunes a huge set of candidate basis functions down to compact models.

@@

	\begin{table}[!h]
	\centering
	\scalebox{0.7}{
		\begin{tabularx}{550pt}{l  l  X}  
			Algorithm          &   Paper/year &   Short details   \\ \hline
			\textsc{GP}        &    \cite{koza1994genetic} (1994)    & Application of GP to SR \\ 
			\textsc{Hybrid-GP}  &   \cite{raidl1998hybrid} (1998)   & GP-based method; solutions are locally optimized with OLS to find optimum coefficients for top-level terms \\ 
			\textsc{DFE}        & \cite{cerny2008using} (2008)   & Differential evolution algorithm \\
			\textsc{APF-FE}                & \cite{schmidt2009distilling, schmidt2011machine} (2009, 2011) & Age-fitness Pareto optimization approach using co-evolved fitness estimation   \\
			APF            &      \cite{schmidt2010age} (2010)        &  Age-fitness Pareto optimization approach                        \\
			\textsc{FFX}  & \cite{mcconaghy2011ffx} (2011)  & The fast function extraction algorithm --  non-evolutionary technique based on a machine learning technique called path-wise regularized learning \\
			\textsc{ABCP} &  \cite{karaboga2012artificial} (2012)  & Artificial bee colony programming  approach \\
			\textsc{EPLEX}          &      \cite{la2016epsilon} (2016)         &   A parent selection method called $\epsilon$-lexicase selection  \\ 
			\textsc{MRGP} & \cite{arnaldo2014multiple} (2014) & It decouples and
			linearly combines a program's subexpressions via multiple regression on the target variable  \\
			Local optimization NLS    & \cite{kommenda2018local} (2018)   & Constants optimization in GP by nonlinear least squares \\
			\textsc{FEAT} & \cite{la2018learning} (2018) & Features are represented as networks of multi-type expression trees comprised of activation functions; differentiable features are trained via gradient descent \\
			\textsc{SBP-GP} &  \cite{virgolin2019linear} (2019)  & The idea of semantic back-propagation utilized in GP \\
			\textsc{BSR} & \cite{jin2019bayesian} (2019) & ML-based approach; Bayesian symbolic regression  \\ 
			\textsc{DSR}  & \cite{petersen2019deep} (2019) & Deep symbolic regression based on a RNN approach further utilizing the policy gradient search \\
			\textsc{Operon} & \cite{kommenda2020parameter} (2020) &  Utilizing nonlinear least squares for parameter identification of SR models with LS \\
			\textsc{VNP}    & \cite{elleuch2020variable} (2020) & VNS based GP approach \\
			\textsc{OccamNet} & \cite{costa2020fast} (2020) &   A fast neural network approach; the model defines a probability distribution over a non-differentiable function space; it samples functions and updates the weights with back-propagation  based on cross-entropy matching in an EA strategy	 \\

			\textsc{AI-Feynman} & \cite{udrescu2020ai} (2020) & A physics-inspired divide-and-conquer method; it also uses neural network fitting \\
			\textsc{GOMEA}  & \cite{virgolin2021improving} (2021)    & A model-based
			EA framework called gene-pool optimal mixing evolutionary algorithm \\
			\textsc{ITEA} & \cite{de2021interaction} (2021)   & EA based approach called the interaction-transformation EA   \\
			\textsc{SA} & \cite{kantor2021simulated} (2021) &  Simulated annealing approach \\

			\textsc{DRLA} & \cite{mundhenk2021symbolic} (2021)  &    A deep reinforcement learning approach enhanced with genetic programming \\
			\textsc{Taylor-GP} &  \cite{he2022taylor} (2022)  &  Taylor polynomials approximations  \\ \hline

	\end{tabularx} }
	\caption{SR methods overview.}
	\label{tab:gp-based}
\end{table}

The SR researches lack uniform, robust, and transparent
benchmarking standards. Recently, La Cava et al.~\cite{la2021contemporary} proposed an
open-source, reproducible benchmarking platform for SR, called \texttt{SRBench}. The authors extended  PMLB~\cite{olson2017pmlb},  a repository of standardized regression problems, with 130 SR datasets for which exact (\emph{ground-truth}) models are known; see more in the aforementioned paper. In their extensive experimental evaluation, 14
symbolic regression methods and 7 machine learning methods are compared on the set of 252 diverse regression problems. The remaining 122 SR datasets belong to a class of so-called \emph{black-box} problems for which the \emph{ground-truth} is unknown. One of the most interesting conclusions from La Cava et al. research is that algorithms specialize in either solving \emph{ground-truth} problems, or \emph{black-box} problems, but not both.

In this work we present a novel approach to solve \emph{ground-truth}-based SR, which combines the popular iterated local search meta-heuristic (ILS)~\cite{lourencco2003iterated,lourencco2010iterated} with the ordinary least square method (OLS)~\cite{leng2007ordinary}. ILS mostly handles combinatorial (discrete) aspects of search space, while OLS helps in the process of a coefficient determination, so it handles some continuous parts of the search space. As will be shown later, the proposed method has shown its robustness w.r.t. introduced noise, therefore we called the method \textsc{RILS}-\textsc{ROLS}  (with letters R corresponding to regression and robust). Additionally, in order to navigate the search toward exact model (and not only accurate one) the algorithm is equipped with a carefully constructed fitness function that combines three important model characteristics: RMSE and $R^2$ scores which participate into models accuracy, and a weighted solution size that penalizes too complex models.

@@

\begin{enumerate}
	\item The proposed method outperforms 14 comparison methods on two \emph{ground-truth} benchmark sets from literature -- these methods and benchmarks are used in \texttt{SRBench} platform.

	\item The method shows its high robustness, which is proved by its comparison to the other algorithms in the presence of different levels of Gaussian white noise in the input data.

	\item The method is very efficient, taking in average less than 4 minutes to reach the exact solution, on the problem instances where it was possible.

	\item A new set of unbiased instances is introduced -- it consists of randomly generated formulae of various sizes and number of input variables. This set was employed to analyze the effect of the model size and the level of noise on the solving difficulty. 
\end{enumerate}

@@

\begin{definition}
	Given is a dataset $D = \{(\mathbf{x_i}, y_i)\}_{i=1}^n$, where $\mathbf{x_i} \in \mathbb{R}^d$ represents i-th feature (input) vector while $y_i \in \mathbb{R}$ is its corresponding target (output) variable. Suppose that there exists an analytical model of the form $f(\mathbf{x})= g^*(\mathbf{x}, \theta^*) + \epsilon $ that is a generator of all observations from $D$.  
	The goal of SR is to learn a mapping $\tilde{f}(\mathbf{x})=  \tilde{g}(\mathbf{x}, \tilde{\theta})  \colon \mathbb{R}^d \mapsto \mathbb{R}$  estimated by searching through the space of (mathematical) expressions  $\tilde{g}$ and parameters $\tilde{\theta}$ where  $\epsilon$ is observed white noise within the given input data.

@@

Koza~\cite{koza1994genetic} introduced the problem of SR as a specific application of genetic programming. GP can be used to optimize nonlinear structures such as computer programs. In particular, the programs are represented by syntax trees consisting of functions/operations over input features and constants. As an example of a function represented by a syntax tree, see Figure~\ref{fig:syntax-tree-example}. In essence, all valid syntax trees form the solution search space of SR. That is, each sample model $\tilde{f}$ may be seen as a point in the search space, represented by respective syntax tree. The solution accuracy can be computed on the basis of historical data $D$ and the chosen error measure such as $MSE$, $RMSE$, $R^2$, some combination of them, etc. Interestingly, the SR search space nature is twofold: discrete and continuous. It is primarily modeled as a problem of discrete (combinatorial) optimization, since the number of possible solution functional forms is countable. However, it may also include elements solved by means of continuous (global) optimization, e.g., constants (coefficients) fitting. It is common to use the set of the following elementary mathematical functions: $\sqrt{x}, x^2 $, $\sin$, $\cos$, $\log$, $\exp$, $\arcsin$, $\arccos$, $a^x$, and a set of standard arithmetic operators: $+$, $-$, $\cdot$, and $/$.

@@

Our method relies on the following operations: $+$, $-$, $\cdot$, $/$, $\sqrt{x}$, $x^2 $, $\sin$, $\cos$, $\log$, $\exp$, $a^x$, where $a^x$ is only consequently used -- it is never explicitly introduced in the search space. Beside this, the following set of constants enters the search space explicitly: $-1$, $0$, $1$, $2$, $\pi$ and $10$. 
Before we give details on our   method for solving SR, we will explain two building blocks of \textsc{RILS-ROLS}: 1) iterated local search (ILS) and 2) ordinary least squares (OLS).

\subsection{Iterated local search}
ILS~\cite{lourencco2003iterated} is an efficient meta-heuristics that iteratively generates a sequence of solutions produced by the (embedded) heuristic, such as local search (LS) or randomized greedy heuristics. When the search gets \emph{stuck} in local optimum, a \emph{perturbation} is performed -- this step is usually non-deterministic. This simple idea was proposed by Baxter~\cite{baxter1981local} in early 1980s, and, since then, has been  re-invented by many researches under different names. Some of the following names were used: iterated descent search~\cite{baum1998iterated}, large-step
Markov chains~\cite{martin1991large}, chained local optimization~\cite{martin1996combining}, or, in some cases, combinations of previous~\cite{applegate2003chained}. The most popular version of ILS is shown in Algorithm~\ref{alg:ils}. (Note that we use this version of ILS as the backbone of our \textsc{RILS}-\textsc{ROLS} algorithm.)

@@

An initial solution may be generated randomly or by using a greedy heuristic, which is afterwards, improved by local search. At each iteration, ILS applies three steps. First, the current incumbent solution $s$ is perturbed -- it is partially randomized, yielding new solution $s'$. Next, the solution $s'$ is potentially improved by a LS procedure. Thirdly, the newly obtained solution $s'$  possibly becomes new incumbent -- this is decided upon the acceptance criterion. Sometimes, ILS incorporates a mechanism of tabu list, which prevents the search getting back into already visited solutions.

>>>>

\title{ {RILS}-{ROLS}: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares}

@@

\begin{abstract} % abstract
In this paper, we solve the well-known symbolic regression problem that has been intensively studied and has a wide range of applications. 
% ACA: ove recenice vise prilice uvodu, tamo svakako i postoji, da ne bude preglomazan abstrakt
%The symbolic regression models could provide insights towards establishing physical theory behind experimentally observed data. 
%Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology, modeling analytic representations of the exciton binding energy in energy science, finding models for band gaps of NaCl-type compounds in material science, just to name a few.

To solve it, we propose an efficient meta-heuristic-based approach, called \textsc{RILS}-\textsc{ROLS}. \textsc{RILS}-\textsc{ROLS} is based on the following two elements: ($i$) iterated local search, which is the method backbone, mainly solving combinatorial and some continuous aspects of the problem; ($ii$) ordinary least square method, which focuses on the continuous aspect of the search space -- it efficiently determines the best--fitting coefficients of linear combinations within solution equations. In addition, we introduce a novel fitness function that combines important model quality measures: $R^2$ score, RMSE score, size of the model (or model complexity), and carefully designed local search, which allows systematic search in proximity to candidate solution.

Experiments are conducted on the two well-known ground-truth benchmark sets called \textsc{Feynman} and \textsc{Strogatz}. \textsc{RILS}-\textsc{ROLS} was compared to 14 other competitors from the literature. Our method outperformed all 14 competitors with respect to the true symbolic model accuracy under varying levels of noise. It also proved to be the most efficient method considering the average running time of reaching the exact model.  
In addition to evaluation on known ground-truth datasets, we introduce a new randomly generated set of problem instances. The goal of the \textsc{Random} dataset was to test the scalability of our method with respect to incremental equation sizes and number of variables, under different levels of noise.

The statistical analysis of obtained experimental results confirmed that \textsc{RILS}-\textsc{ROLS} can be considered as a new state-of-the-art method for solving the problem of symbolic regression when ground-truth equations are known.
\end{abstract}

@@

The problem of symbolic regression (SR)~\cite{billard2002symbolic} has attracted a lot of research interest over the last decade. SR can be seen as a generalization of more specific variants of regression in which the functional form is fixed, e.g., the well-known linear regression, polynomial regression~\cite{stimson1978interpreting}, etc. All regression models have the same goal: given a set of $n$-dimensional input data and its corresponding continuous output target variable, the aim is to find a  mathematical expression (function) of $n$ (input) variables that best \emph{fits} the target variable.  %This computationally intensive task is in general provenly NP--hard~\cite{virgolin2022symbolic}. 
In the case of linear regression, the model is always a linear combination of input variables. This is in general not good enough, since the target variable might be dependent on a nonlinear function among input variables. Unlike linear regression, SR allows for the search over much larger space of possible mathematical formulas to find the best-fitting ones, i.e., those able to predict the target variable from input variables. The basis of constructing an explicit formula is in elementary operations like addition and multiplication, as well as polynomial, trigonometric, exponential, and other operations.  
%Application: https://towardsdatascience.com/real-world-applications-of-symbolic-regression-2025d17b88ef

SR models tend to be interpretable, unlike artificial neural networks. Coefficients inside SR formulas might indicate the absolute or relative importance of certain input variables. Moreover, relations among variables might be described in an elegant way. For example, the appearance of an exponential function can be associated with a specific physical phenomenon such as the intensity of radiation or acceleration over time (see~\cite{udrescu2020ai}). Additionally, SR models often have high generalization power unlike some models with fixed functional forms such as polynomial regression.

Practical applications of SR in chemical and biological sciences are listed in~\cite{weng2020simple}. This paper in particular describes the discovery of a series of new oxide perovskite catalysts with improved activities. The application of SR to discovering physical laws from a distorted video is studied in~\cite{udrescu2021symbolic}. The revealing complex ecological dynamics by SR is presented in~\cite{chen2019revealing}. The paper \cite{louis2021reviewing} presents the application of the SR to model the effects of mutations on protein stability in the domain of fundamental and applied biology. A recent study in \cite{liu2021machine} is concerned with auto-discovering conserved quantities using trajectory data from unknown dynamical systems. The application of the SR to model analytic representations of exciton binding energy is shown in~\cite{liang2019phillips}. The use of SR in material science is described in~\cite{wang2019symbolic,wang2022symbolic,burlacu2022symbolic,kabliman2021application}. SR application to wind speed forecasting is given in~\cite{abdellaoui2021symbolic}.

There are many different ways to tackle SR. Most of them are based on   machine learning techniques, genetic programming (GP), or some other meta-heuristics. Among the first GP methods to tackle SR is the one of Raidl~\cite{raidl1998hybrid} based on a hybrid variant of genetic programming. A differential evolution algorithm is proposed by Cerny et al.~\cite{cerny2008using}. Age-fitness Pareto Optimization approach is proposed by Smidt and Lipson~\cite{schmidt2010age}.  Application of the artificial bee colony programming to solve SR is proposed by Karaboga et al.~\cite{karaboga2012artificial}. 
Application of local-based heuristics for solving SR is reported by Commenda in his PhD thesis~\cite{kommenda2018local}. A GP-based approach, the gene-pool optimal mixing evolutionary algorithm (\textsc{GOMEA}) is studied by Virgolin et al.~\cite{virgolin2021improving}.  Another evolutionary algorithm, the interaction-transformation EA (\textsc{ITEA}) has been proposed by de Franca et al.~\cite{de2021interaction}. Simulated annealing to solve SR is proposed by Kantor~\cite{kantor2021simulated}. A variable neighbourhood programming approach  is proposed by Elleurich et al.~\cite{elleuch2020variable}; this technique was initially proposed in~\cite{elleuch2016variable}. 
Kommenda et al.~\cite{kommenda2020parameter} proposed a method called \textsc{Operon} algorithm, which uses nonlinear least squares for parameter identification of SR models further integrated into a local search mechanism in tree-based GP. The C++ implementation of \textsc{Operon} is discussed in~\cite{burlacu2020operon}. The method that utilizes Taylor polynomials to approximate the symbolic equation that fits the dataset, called Taylor genetic programming, is proposed in~\cite{he2022taylor}. A GP approach that uses the idea of semantic back-propagation (SBP-GP) is proposed in~\cite{virgolin2019linear}.  The empirical analysis of variance between many GP-based methods for SR is discussed in~\cite{kammerer2021empirical}. It is also worth to mention the GP-based Eurequa commercial solver~\cite{schmidt2009distilling, schmidt2011machine} that uses age-fitness pareto optimization with co-evolved fitness estimation. This solver is nowadays accepted as the gold standard of symbolic regression.

The method based on Bayesian symbolic regression (\textsc{BSR}) is proposed in~\cite{jin2019bayesian} -- this method belongs to the family of Markov Chain Monte Carlo algorithms (MCMC). Deep Symbolic Regression (\textsc{DSR}), an RNN approach that utilizes the policy gradient search, is proposed in~\cite{petersen2019deep}. This mechanism of search is further investigated in~\cite{landajuela2021improving}. A fast neural network approach, called \textsc{OccamNet} is proposed in~\cite{costa2020fast}.  A deep reinforcement learning approach enhanced with genetic programming is proposed in~\cite{mundhenk2021symbolic}.

Powerful hybrid techniques for solving SR are also well studied; among them, we emphasize the \textsc{Eplex} solver by~\cite{la2019probabilistic,la2016epsilon} and \textsc{AI-Feynman} algorithm by~\cite{udrescu2020ai}, based on a physics-inspired divide-and-conquer method combined with the neural network fitting. The latter is one of the most efficient methods for physically-inspired models. We would also like to mention the Fast Function Extraction (\textsc{FFX}) algorithm developed by McConaghy~\cite{mcconaghy2011ffx}, which is a non-evolutionary method combined with a machine learning technique called path-wise-regularized learning, which quickly prunes a huge set of candidate basis functions down to compact models.

@@

	\begin{table}[!h]
		\caption{SR methods overview.}
		\label{tab:gp-based}
	\centering
	\scalebox{0.7}{
		\begin{tabularx}{550pt}{l  l  X}  
			Algorithm          &   Paper/year &   Short details   \\ \hline
			\textsc{GP}        &    \cite{koza1994genetic} (1994)    & Application of GP to SR. \\ 
			\textsc{Hybrid-GP}  &   \cite{raidl1998hybrid} (1998)   & GP-based method; solutions are locally optimized with OLS to find optimal coefficients for top-level terms. \\ 
			\textsc{DFE}        & \cite{cerny2008using} (2008)   & Differential evolution algorithm. \\
			\textsc{APF-FE}                & \cite{schmidt2009distilling, schmidt2011machine} (2009, 2011) & Age-fitness Pareto optimization approach using co-evolved fitness estimation.   \\
			\textsc{APF}            &      \cite{schmidt2010age} (2010)        &  Age-fitness Pareto optimization approach.                        \\
			\textsc{FFX}  & \cite{mcconaghy2011ffx} (2011)  & The fast function extraction algorithm --  non-evolutionary technique based on a machine learning technique called path-wise regularized learning. \\
			\textsc{ABCP} &  \cite{karaboga2012artificial} (2012)  & Artificial bee colony programming  approach. \\
			\textsc{EPLEX}          &      \cite{la2016epsilon} (2016)         &   A parent selection method called $\epsilon$-lexicase selection.  \\ 
			\textsc{MRGP} & \cite{arnaldo2014multiple} (2014) & It decouples and
			linearly combines a program's subexpressions via multiple regression on the target variable.  \\
			Local optimization NLS    & \cite{kommenda2018local} (2018)   & Constants optimization in GP by nonlinear least squares. \\
			\textsc{FEAT} & \cite{la2018learning} (2018) & Features are represented as networks of multi-type expression trees comprised of activation functions; differentiable features are trained via gradient descent. \\
			\textsc{SBP-GP} &  \cite{virgolin2019linear} (2019)  & The idea of semantic back-propagation utilized in GP. \\
			\textsc{BSR} & \cite{jin2019bayesian} (2019) & ML-based approach; Bayesian symbolic regression. \\ 
			\textsc{DSR}  & \cite{petersen2019deep} (2019) & Deep symbolic regression based on a RNN approach further utilizing the policy gradient search. \\
			\textsc{Operon} & \cite{kommenda2020parameter} (2020) &  Utilizing nonlinear least squares for parameter identification of SR models with LS. \\
			\textsc{VNP}    & \cite{elleuch2020variable} (2020) & VNS approach inspired by GP. \\
			\textsc{OccamNet} & \cite{costa2020fast} (2020) &   A fast neural network approach; the model defines a probability distribution over a non-differentiable function space; it samples functions and updates the weights with back-propagation based on cross-entropy matching in an EA strategy.	 \\

			\textsc{AI-Feynman} & \cite{udrescu2020ai} (2020) & A physics-inspired divide-and-conquer method; it also uses neural network fitting. \\
			\textsc{GOMEA}  & \cite{virgolin2021improving} (2021)    & A model-based
			EA framework called gene-pool optimal mixing evolutionary algorithm. \\
			\textsc{ITEA} & \cite{de2021interaction} (2021)   & EA based approach called the interaction-transformation EA.   \\
			\textsc{SA} & \cite{kantor2021simulated} (2021) &  Simulated annealing approach. \\

			\textsc{DRLA} & \cite{mundhenk2021symbolic} (2021)  &    A deep reinforcement learning approach enhanced with genetic programming. \\
			\textsc{Taylor-GP} &  \cite{he2022taylor} (2022)  &  Taylor polynomials approximations.  \\ \hline

	\end{tabularx} }
\end{table}

SR research lacks uniform, robust, and transparent
benchmarking standards. Recently, La Cava et al.~\cite{la2021contemporary} proposed an
open-source, reproducible benchmarking platform for SR called \texttt{SRBench}. The authors extended  PMLB~\cite{olson2017pmlb},  a repository of standardized regression problems, with 130 SR datasets for which exact (\emph{ground-truth}) models are known; see more in the aforementioned paper. In their extensive experimental evaluation, 14
symbolic regression methods and 7 machine learning methods are compared on the set of 252 diverse regression problems. The remaining 122 SR datasets belong to a class of so-called \emph{black-box} problems for which the ground-truth is unknown. One of the most interesting conclusions from La Cava et al. research is that algorithms specialize in either solving ground-truth problems, or black-box problems, but not both.

In this work we present a novel approach to solving ground-truth-based SR, which combines the popular iterated local search meta-heuristic (ILS)~\cite{lourencco2003iterated,lourencco2010iterated} with the ordinary least square method (OLS)~\cite{leng2007ordinary}. ILS mostly handles combinatorial (discrete) aspects of search space, while OLS helps in the process of coefficient determination, so it handles some continuous parts of the search space. As will be shown later, the proposed method has shown its robustness w.r.t. introduced noise, which is why we called the method \textsc{RILS}-\textsc{ROLS}  (with letters R for regression and robust). Additionally, to navigate the search toward the exact model (and not only the accurate one) the algorithm is equipped with a carefully constructed fitness function that combines three important model characteristics: RMSE and $R^2$ scores which participate int the model's accuracy, and a weighted solution size that penalizes too complex models.

@@

\begin{enumerate}
	\item The proposed method outperforms 14 comparison methods on two ground-truth benchmark sets from the literature -- these methods and benchmarks are used in the \texttt{SRBench} platform.

	\item The method demonstrates high robustness, which is proved by comparison to other algorithms in the presence of different levels of Gaussian white noise in data.

	\item The method is very efficient, taking on average less than 4 minutes to reach the exact solution in the problem instances where it was possible.

	\item A new set of unbiased instances is introduced -- it consists of randomly generated formulae of various sizes and a number of input variables. This set was employed to analyze the effect of model size, number of input variables and the level of noise on the solving difficulty. 
\end{enumerate}

@@

\begin{definition}
	Given is a dataset $D = \{(\mathbf{x_i}, y_i)\}_{i=1}^n$, where $\mathbf{x_i} \in \mathbb{R}^d$ represents the i-th feature (input) vector while $y_i \in \mathbb{R}$ is its corresponding target (output) variable. Suppose that there exists an analytical model of the form $f(\mathbf{x})= g^*(\mathbf{x}, \theta^*) + \epsilon $ that is a generator of all observations from $D$.  
	The goal of SR is to learn the mapping $\tilde{f}(\mathbf{x})=  \tilde{g}(\mathbf{x}, \tilde{\theta})  \colon \mathbb{R}^d \mapsto \mathbb{R}$  estimated by searching through the space of (mathematical) expressions  $\tilde{g}$ and parameters $\tilde{\theta}$ where  $\epsilon$ is the observed white noise within the given input data.

@@

Koza~\cite{koza1994genetic} introduced the problem of SR as a specific application of genetic programming. GP can be used to optimize nonlinear structures such as computer programs. In particular, programs are represented by syntax trees consisting of functions/operations over input features and constants. As an example of a function represented by a syntax tree, see Figure~\ref{fig:syntax-tree-example}. In essence, all valid syntax trees form the solution search space of SR. That is, each sample model $\tilde{f}$ may be seen as a point in the search space, represented by a respective syntax tree. The solution accuracy can be computed on the basis of historical data $D$ and the chosen error measure such as $MSE$, $RMSE$, $R^2$, their combination, etc. Interestingly, the nature of SR search space is twofold: discrete and continuous. It is primarily modeled as a problem of discrete (combinatorial) optimization, since the number of possible solution functional forms is countable. However, it may also include elements solved by means of continuous (global) optimization, e.g., constants (coefficients) fitting. It is common to use the set of the following elementary mathematical functions: $\sqrt{x}, x^2 $, $\sin$, $\cos$, $\log$, $\exp$, $\arcsin$, $\arccos$, $a^x$, and a set of standard arithmetic operators: $+$, $-$, $\cdot$, and $/$.

@@

Our method relies on the following operations: $+$, $-$, $\cdot$, $/$, $\sqrt{x}$, $x^2 $, $\sin$, $\cos$, $\log$, $\exp$, $a^x$, where $a^x$ is only consequently used -- it is never explicitly introduced in the search space. Beside these, the following set of constants enters the search space explicitly: $-1$, $0$, $1$, $2$, $\pi$, and $10$. 
Before we provide details of our method for solving SR, we will explain the two building blocks of \textsc{RILS-ROLS}: 1) iterated local search (ILS) and 2) ordinary least squares (OLS).

\subsection{Iterated local search}
ILS~\cite{lourencco2003iterated} is an efficient meta-heuristic that iteratively generates a sequence of solutions produced by the (embedded) heuristic, such as local search (LS) or randomized greedy heuristics. When the search gets \emph{stuck} in the local optimum, \emph{perturbation} is performed, which is usually a non-deterministic step. This simple idea was proposed by Baxter~\cite{baxter1981local} in the early 1980s, and has since then been re-invented by many researchers under different names. These are some of the names that were used: iterated descent search~\cite{baum1998iterated}, large-step
Markov chains~\cite{martin1991large}, chained local optimization~\cite{martin1996combining}, or in some cases, combinations of those~\cite{applegate2003chained}. The most popular version of ILS is shown in Algorithm~\ref{alg:ils}. (Note that we use this version of ILS as the backbone of our \textsc{RILS}-\textsc{ROLS} algorithm.)

@@

An initial solution may be generated randomly or by using a greedy heuristic, which is afterwards improved by local search. At each iteration, ILS applies three steps. First, the current incumbent solution $s$ is perturbed, i.e., partially randomized, yielding a new solution $s'$. Next, the solution $s'$ is potentially improved by an LS procedure. Thirdly, the newly obtained solution $s'$  possibly becomes a new incumbent -- this is decided upon the acceptance criterion. Sometimes, ILS incorporates the mechanism of a tabu list, which prevents the search from getting back into already visited solutions.

\subsection{Ordinary least square method}\label{sec:ols}
The ordinary least square method (OLS) is a linear regression technique. It is based on applying the least-square method to minimize the square residual (error) sum  between actual and predicted values (given by the model). More precisely, given the   dataset $D$ of $n$ points $(\mathbf{x_i}, y_i)$ where each $\mathbf{x_i}$ is $d$-dimensional, the task is to determine linear mapping $\tilde{y} = \mathbf{k} \mathbf{x} + b$, that is coefficients (line slope) $\mathbf{k} = (k_1, \ldots, k_d)$ and $b$ (intercept), so that $ \sum_{i}^{n} (\tilde{y}_i - y_i)^2 $ is minimized. This sum is also known as the sum of squared errors (SSE). There are many methods to minimize SSE. One of the analytical approaches is calculus-oriented so it takes into account the partial derivatives of SSE w.r.t. $k_j,\; j \in \{1, ..., d\}$:
